---
title: "Your Turn Solutions"
author: "Heike Hofmann"
date: "May, 2018"
output: html_document
---

## Graphics Intro

Load the BRFSS subset for Iowa in 2012

```{r}
iowa <- read.csv("https://raw.githubusercontent.com/heike/summerschool-2018/master/02-graphics/data/brfss-iowa-2012.csv")
```

1. Use `qplot` to build a scatterplot of variables `WTKG3` (weight in kg) and `HTIN4` (height in inch)

```{r}
library(ggplot2)
qplot(data = iowa, x = HTIN4, y = WTKG3)

# use some alpha blending to alleviate over-plotting
qplot(data = iowa, x = HTIN4, y = WTKG3, alpha = I(0.5))

# Generally there is a positive relationship, but it's not very strong, and there are lots of outliers on the upper end of the weight.
# The stripes come from height being reported in inch. We can use jittering to add some random noise

qplot(data = iowa, x = HTIN4, y = WTKG3, alpha = I(0.5), geom="jitter")

```
2. Use options within `qplot` to color points by a person's gender (`SEX`)

```{r}
qplot(data = iowa, x = HTIN4, y = WTKG3/100, colour = SEX)

# the colour is strange - let's change the SEX variable to a factor permanently:
iowa$SEX <- factor(iowa$SEX)

qplot(data=iowa, x = HTIN4, y = WTKG3, colour = SEX)
```

3. Clean up axis labels and add main plot title

```{r}
qplot(data = iowa, x = HTIN4, y = WTKG3, colour = SEX) +
  ggtitle("Relationship of weight and height among the Iowa population in 2012") +
  xlab("Height in inch") +
  ylab("Weight in kg")
```

## boxplots

1. Create a new variable in the brfss data calculating bmi: bmi = (weight in kg)^2 / height in cm

```{r}
iowa$bmi <- (iowa$WTKG3/100)/(iowa$HTIN4*2.54/100)^2
summary(iowa$bmi)
```
2. Use qplot to create a histogram of bmi
```{r}
qplot(data = iowa, x = bmi, geom="histogram") 
```

3. Change the bin width on that histogram to 0.5

```{r}
qplot(data = iowa, x = bmi, geom="histogram", binwidth=1) 
```

4. Facet this histogram by gender

```{r}
qplot(data = iowa, x = bmi, geom="histogram", facets = SEX~.)
```

## Binge Drinking

1. The variable `DRNK3GE5` captures the number of times in the last 30 days that respondents binge drank. Draw side-by-side boxplots of  `DRNK3GE5` for genders (`SEX`).

```{r}
qplot(data = iowa, x = SEX, y = DRNK3GE5, geom="boxplot") 
```

2. It turns out that values 88, 99 and 77 have special meaning: 88 stands for "none", 99 is "refused to answer" and 77 is "not sure". Exclude these values from the boxplot.

```{r}
qplot(data = subset(iowa, DRNK3GE5<77), x = SEX, y = DRNK3GE5, geom="boxplot") 
```

3. Overlay jittered points for observed values onto this boxplot

```{r}
qplot(data = subset(iowa, DRNK3GE5<77), x = SEX, y = DRNK3GE5, geom="boxplot") +
  geom_jitter()
```

## Seatbelt use

1. Use the brfss data to make a barplot for counts of seatbelt use (SEATBELT).
*Seatbelt use (SEATBELT) is captured as 1 - Always, 2- Nearly Always, 3-Sometimes, 4- Seldom, 5-Never, 7-Don't know, 8-Never drive or ride in a car, 9-Refused, to BLANK-Missing.*

```{r}
qplot(data = iowa, x = SEATBELT, geom="bar")
qplot(data = iowa, x = factor(SEATBELT), geom="bar")
```

2. Facet seatbelt use by gender - are there gender differences in usage?

```{r}
qplot(data = iowa, x = factor(SEATBELT), geom="bar", facets=~SEX)
# proportion of women always using a seatbelt is higher than proportion of men.
qplot(data = iowa, x = factor(SEATBELT), geom="bar", fill=factor(SEX))

qplot(data = iowa, x = SEX, geom="bar", fill=SEATBELT==1)

```
